PANCHANGE_data_preprocessing

### 2020-08-07 PANCHANGE_data_preprocessing2020-07-15.Rmd
cleaning respiratory

```{r glad phq cleaning}
glad.phq.raw.id <- glad.phq.raw %>%
  drop_na(externalDataReference) %>% # Drop NAs
  distinct(externalDataReference, .keep_all = TRUE) %>% # Remove duplicates based on ID
  separate(externalDataReference, into = c("Sample", "ID"), sep = 4) %>% # Split ID in Sample and Number
  mutate(ID = as.numeric(ID),
         startDate.glad = startDate) %>% # Rename startDate
  select(
    Sample, # Sample
    ID, # ID
    phq.problems_made_difficult_care, # Impairment
    phq.pandemic_felt_feelings, # Comparison to before the pandemic (retro)
    phq.little_interest_or_pleasure_in_doing_things_prepan = phq.little_interest_or_pleasure_in_doing_things,
    phq.feeling_down_depressed_or_hopeless_prepan = phq.feeling_down_depressed_or_hopeless,
    phq.staying_asleep_sleeping_trouble_prepan = phq.staying_asleep_sleeping_trouble,
    phq.feeling_tired_or_having_little_energy_prepan = phq.feeling_tired_or_having_little_energy,
    phq.poor_appetite_or_overeating_prepan = phq.poor_appetite_or_overeating,
    phq.feeling_bad_failure_family_prepan = phq.feeling_bad_failure_family,
    phq.trouble_concentrating_reading_newspaper_prepan = phq.trouble_concentrating_reading_newspaper,
    phq.moving_fidgety_noticed_opposite_prepan = phq.moving_fidgety_noticed_opposite,
    phq.dead_hurting_thoughts_prepan = phq.dead_hurting_thoughts
  ) %>%
  add_numeric(., exclude = exclude_cols) %>%
  mutate_if(is.numeric, ~na_if(., -88)) %>% # Recode missing values to NAs in numeric variables
  mutate_if(is.numeric, ~na_if(., -99)) %>%
  mutate_if(is.numeric, ~na_if(., -77)) %>%
  mutate_if(is.factor, ~na_if(., "Seen but not answered")) %>% # Recode missing values to NAs in factor variables
  mutate_if(is.factor, ~na_if(., "Don't know")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to say")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Seen but not answered")) %>% # Drop empty factor levels
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Don't know")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to say"))

@@

    ocir.felt_pandemic_feelings,

@@

Select respiratory variables
```{r read in respiratory data}
coping.glad.respiratory.raw.id <- coping.glad.respiratory.raw %>%
  drop_na(externalDataReference) %>% # Drop NAs
  distinct(externalDataReference, .keep_all = TRUE) %>% # Remove duplicates based on ID
  separate(externalDataReference, into = c("Sample", "ID"), sep = 4) %>% # Split ID in Sample and Number
  mutate(ID = as.numeric(ID),
         startDate.glad = startDate) %>% # Rename startDate
  select(
    Sample, # Sample
    ID,
    respiratory.breath_diarrhoea_fatigue_fever,
    respiratory.worst_ill_feel,
    respiratory.persistent_cough,
    respiratory.skipping_meals_because_you_felt_unwell,
    respiratory.was_your_sense_of_smell_impaired,
    respiratory.was_your_sense_of_taste_impaired,
    respiratory.feeling_tiredfatigued,
    respiratory.throat_swab_coronavirus_nose,
    respiratory.throat_swab_test_nose,
    respiratory.coronavirus_antibody_test,
    respiratory.results_antibody_test
    # ID
  ) %>%
  mutate_if(is.numeric, ~na_if(., -88)) %>% # Recode missing values to NAs in numeric variables
  mutate_if(is.numeric, ~na_if(., -99)) %>%
  mutate_if(is.numeric, ~na_if(., -77)) %>%
  mutate_if(is.factor, ~na_if(., "Seen but not answered")) %>% # Recode missing values to NAs in factor variables
  mutate_if(is.factor, ~na_if(., "Don't know")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to say")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Seen but not answered")) %>% # Drop empty factor levels
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Don't know")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to say"))

@@

```{r create respiratory numeric variables}
exclude_cols <- c("ID",
                  "Sample")

coping.glad.respiratory.raw.numeric <- add_numeric(coping.glad.respiratory.raw.id, exclude = exclude_cols)

colnames(coping.glad.respiratory.raw.numeric)
dim(coping.glad.respiratory.raw.numeric)
dim(coping.glad.respiratory.raw.id)
```

#Distribution of likert scale COVID-19 symptoms in COPING
```{r distribution of COVID symptoms}
#persistent cough
coping.glad.respiratory.raw.numeric %>%
  ggplot(
    mapping = aes(
      y = respiratory.persistent_cough_numeric 
    )
  ) +
  geom_bar() +
  labs(
    title = "Persistant cough"
  )

#severe fatigue
coping.glad.respiratory.raw.numeric %>%
  ggplot(
    mapping = aes(
        y = respiratory.feeling_tiredfatigued_numeric 
    )
  ) +
  geom_bar() +
  labs(
    title = "Severe fatigue"
  )

#skipped meals
coping.glad.respiratory.raw.numeric %>%
  ggplot(
    mapping = aes(
        y = respiratory.skipping_meals_because_you_felt_unwell_numeric
    )
  ) +
  geom_bar() +
  labs(
    title = "Skipped meals"
  )

@@

Results:
Persistent cough: "Mild symptoms" as cut off
Fatigue: "Severe symptoms" as cut off
Skipped meals: "Moderate symptoms" as cut off
```{r proportion of people who report COVID symptoms in Menni et al}
#persistent cough
percentage_pcough <- ((0.5673*6452) + (0.4556*9186) + (0.3017*2434931))/(6452+9186+2434931)
percentage_pcough    # 0.3029762% of Menni et al(2020) have fatigue
quantile(x = coping.glad.respiratory.raw.numeric$respiratory.persistent_cough_numeric, 
         na.rm = TRUE, 
         probs = 1-percentage_pcough) # this gives 0.3029762% of the sample reporting fatigue when using the "mild symptoms" persistent cough response

#fatigue
percentage_fatigue <- ((0.298*6452) + (0.1556*9186) + (0.0871*2434931))/(6452+9186+2434931) 
percentage_fatigue   # 0.08791204% of Menni et al(2020) have fatigue
quantile(x = coping.glad.respiratory.raw.numeric$respiratory.feeling_tiredfatigued_numeric, 
         na.rm = TRUE, 
         probs = 1-percentage_fatigue) #this gives 8% of the sample reporting fatigue when using the "severe symptoms" fatigue response

#skipped meals 
percentage_skipped_meals <- ((0.4203*6452) + (0.2493*9186) + (0.1912*2434931))/(6452+9186+2434931) 
percentage_skipped_meals    # 0.192021% of Menni et al(2020) have fatigue
quantile(x = coping.glad.respiratory.raw.numeric$respiratory.skipping_meals_because_you_felt_unwell_numeric, 
         na.rm = TRUE, 
         probs = 1-percentage_skipped_meals) # this gives 0.192021% of the sample reporting fatigue when using the "moderate symptoms" persistent cough response
```

Creating binary groups for COVID symptoms
```{r creating binary groups for COVID symptoms}
#loss or smell or taste
coping.glad.respiratory.raw.numeric <- coping.glad.respiratory.raw.numeric %>%
  mutate(
    respiratory.anosmia_binary =
      if_else(
        respiratory.was_your_sense_of_smell_impaired_numeric == 1 | respiratory.was_your_sense_of_taste_impaired_numeric == 1, 
        1,
        0
      )
  )

#persistent cough
coping.glad.respiratory.raw.numeric <- coping.glad.respiratory.raw.numeric %>%
  mutate(
    respiratory.pcough_binary =
      if_else(
        respiratory.persistent_cough_numeric >= 1, #those who score 1 or above (mild or above)
        1,
        0
      )
  )

#fatigue
coping.glad.respiratory.raw.numeric <- coping.glad.respiratory.raw.numeric %>%
  mutate(
    respiratory.fatigue_binary =
      if_else(
        respiratory.feeling_tiredfatigued_numeric >= 3, #those who score 3 (severe symptoms)
        1,
        0
      )
  )

#skipped meals
coping.glad.respiratory.raw.numeric <- coping.glad.respiratory.raw.numeric %>%
  mutate(
    respiratory.skipped_meals_binary =
      if_else(
        respiratory.skipping_meals_because_you_felt_unwell_numeric >= 2,
        1,
        0
      )
  )
```

Frequencies of binary COVID symptoms
```{r frequencies of COVID binary symptoms}
freq(coping.glad.respiratory.raw.numeric$respiratory.anosmia_binary)
freq(coping.glad.respiratory.raw.numeric$respiratory.pcough_binary)
freq(coping.glad.respiratory.raw.numeric$respiratory.fatigue_binary)
freq(coping.glad.respiratory.raw.numeric$respiratory.skipped_meals_binary)
```

Rename respiratory dataset joined to main dataset
```{r clean respiratory dataset}
coping.glad.respiratory <- coping.glad.respiratory.raw.numeric

>>>>

```{r glad phq cleaning}
glad.phq.raw.id <- glad.phq.raw %>%
  drop_na(externalDataReference) %>% # Drop NAs
  distinct(externalDataReference, .keep_all = TRUE) %>% # Remove duplicates based on ID
  separate(externalDataReference, into = c("Sample", "ID"), sep = 4) %>% # Split ID in Sample and Number
  mutate(ID = as.numeric(ID),
         startDate.glad = startDate) %>% # Rename startDate
  select(
    Sample, # Sample
    ID, # ID
    phq.little_interest_or_pleasure_in_doing_things_prepan = phq.little_interest_or_pleasure_in_doing_things,
    phq.feeling_down_depressed_or_hopeless_prepan = phq.feeling_down_depressed_or_hopeless,
    phq.staying_asleep_sleeping_trouble_prepan = phq.staying_asleep_sleeping_trouble,
    phq.feeling_tired_or_having_little_energy_prepan = phq.feeling_tired_or_having_little_energy,
    phq.poor_appetite_or_overeating_prepan = phq.poor_appetite_or_overeating,
    phq.feeling_bad_failure_family_prepan = phq.feeling_bad_failure_family,
    phq.trouble_concentrating_reading_newspaper_prepan = phq.trouble_concentrating_reading_newspaper,
    phq.moving_fidgety_noticed_opposite_prepan = phq.moving_fidgety_noticed_opposite,
    phq.dead_hurting_thoughts_prepan = phq.dead_hurting_thoughts
  ) %>%
  add_numeric(., exclude = exclude_cols) %>%
  mutate_if(is.numeric, ~na_if(., -88)) %>% # Recode missing values to NAs in numeric variables
  mutate_if(is.numeric, ~na_if(., -99)) %>%
  mutate_if(is.numeric, ~na_if(., -77)) %>%
  mutate_if(is.factor, ~na_if(., "Seen but not answered")) %>% # Recode missing values to NAs in factor variables
  mutate_if(is.factor, ~na_if(., "Don't know")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to say")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Seen but not answered")) %>% # Drop empty factor levels
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Don't know")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to say"))

@@

    ocir.felt_pandemic_feelings, # Comparison to before the pandemic

@@

Select respiratory variables
```{r read in respiratory data}
coping.glad.respiratory.raw.id <- coping.glad.respiratory.raw %>%
  drop_na(externalDataReference) %>% # Drop NAs
  distinct(externalDataReference, .keep_all = TRUE) %>% # Remove duplicates based on ID
  separate(externalDataReference, into = c("Sample", "ID"), sep = 4) %>% # Split ID in Sample and Number
  mutate(ID = as.numeric(ID),
         startDate.glad = startDate) %>% # Rename startDate
  select(
    Sample, # Sample
    ID, # ID
    respiratory.breath_diarrhoea_fatigue_fever,
    respiratory.worst_ill_feel,
    respiratory.persistent_cough,
    respiratory.skipping_meals_because_you_felt_unwell,
    respiratory.was_your_sense_of_smell_impaired,
    respiratory.was_your_sense_of_taste_impaired,
    respiratory.feeling_tiredfatigued,
    respiratory.throat_swab_coronavirus_nose,
    respiratory.throat_swab_test_nose,
    respiratory.coronavirus_antibody_test,
    respiratory.results_antibody_test
  ) %>%
  add_numeric(., exclude = exclude_cols) %>%
  mutate_if(is.numeric, ~na_if(., -88)) %>% # Recode missing values to NAs in numeric variables
  mutate_if(is.numeric, ~na_if(., -99)) %>%
  mutate_if(is.numeric, ~na_if(., -77)) %>%
  mutate_if(is.factor, ~na_if(., "Seen but not answered")) %>% # Recode missing values to NAs in factor variables
  mutate_if(is.factor, ~na_if(., "Don't know")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to say")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Seen but not answered")) %>% # Drop empty factor levels
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Don't know")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to say"))

@@

#Distribution of likert scale COVID-19 symptoms in COPING
```{r distribution of COVID symptoms}
#persistent cough
coping.glad.respiratory.raw.id %>%
  ggplot(
    mapping = aes(
      y = respiratory.persistent_cough_numeric 
    )
  ) +
  geom_bar() +
  labs(
    title = "Persistant cough"
  )

#severe fatigue
coping.glad.respiratory.raw.id %>%
  ggplot(
    mapping = aes(
        y = respiratory.feeling_tiredfatigued_numeric 
    )
  ) +
  geom_bar() +
  labs(
    title = "Severe fatigue"
  )

#skipped meals
coping.glad.respiratory.raw.id %>%
  ggplot(
    mapping = aes(
        y = respiratory.skipping_meals_because_you_felt_unwell_numeric
    )
  ) +
  geom_bar() +
  labs(
    title = "Skipped meals"
  )

@@

Results:
Persistent cough: "Mild symptoms" as cut off
Fatigue: "Severe symptoms" as cut off
Skipped meals: "Moderate symptoms" as cut off
```{r proportion of people who report COVID symptoms in Menni et al}
#persistent cough
percentage_pcough <- ((0.5673*6452) + (0.4556*9186) + (0.3017*2434931))/(6452+9186+2434931)
percentage_pcough    # 0.3029762% of Menni et al(2020) have fatigue
quantile(x = coping.glad.respiratory.raw.id$respiratory.persistent_cough_numeric, 
         na.rm = TRUE, 
         probs = 1-percentage_pcough) # this gives 0.3029762% of the sample reporting fatigue when using the "mild symptoms" persistent cough response

#fatigue
percentage_fatigue <- ((0.298*6452) + (0.1556*9186) + (0.0871*2434931))/(6452+9186+2434931) 
percentage_fatigue   # 0.08791204% of Menni et al(2020) have fatigue
quantile(x = coping.glad.respiratory.raw.id$respiratory.feeling_tiredfatigued_numeric, 
         na.rm = TRUE, 
         probs = 1-percentage_fatigue) #this gives 8% of the sample reporting fatigue when using the "severe symptoms" fatigue response

#skipped meals 
percentage_skipped_meals <- ((0.4203*6452) + (0.2493*9186) + (0.1912*2434931))/(6452+9186+2434931) 
percentage_skipped_meals    # 0.192021% of Menni et al(2020) have fatigue
quantile(x = coping.glad.respiratory.raw.id$respiratory.skipping_meals_because_you_felt_unwell_numeric, 
         na.rm = TRUE, 
         probs = 1-percentage_skipped_meals) # this gives 0.192021% of the sample reporting fatigue when using the "moderate symptoms" persistent cough response
```

Creating binary groups for COVID symptoms
```{r creating binary groups for COVID symptoms}
#loss or smell or taste
coping.glad.respiratory.raw.id <- coping.glad.respiratory.raw.id %>%
  mutate(
    respiratory.anosmia_binary =
      if_else(
        respiratory.was_your_sense_of_smell_impaired_numeric == 1 | respiratory.was_your_sense_of_taste_impaired_numeric == 1, 
        1,
        0
      )
  )

#persistent cough
coping.glad.respiratory.raw.id <- coping.glad.respiratory.raw.id %>%
  mutate(
    respiratory.pcough_binary =
      if_else(
        respiratory.persistent_cough_numeric >= 1, #those who score 1 or above (mild or above)
        1,
        0
      )
  )

#fatigue
coping.glad.respiratory.raw.id <- coping.glad.respiratory.raw.id %>%
  mutate(
    respiratory.fatigue_binary =
      if_else(
        respiratory.feeling_tiredfatigued_numeric >= 3, #those who score 3 (severe symptoms)
        1,
        0
      )
  )

#skipped meals
coping.glad.respiratory.raw.id <- coping.glad.respiratory.raw.id %>%
  mutate(
    respiratory.skipped_meals_binary =
      if_else(
        respiratory.skipping_meals_because_you_felt_unwell_numeric >= 2,
        1,
        0
      )
  )
```

Frequencies of binary COVID symptoms
```{r frequencies of COVID binary symptoms}
freq(coping.glad.respiratory.raw.id$respiratory.anosmia_binary)
freq(coping.glad.respiratory.raw.id$respiratory.pcough_binary)
freq(coping.glad.respiratory.raw.id$respiratory.fatigue_binary)
freq(coping.glad.respiratory.raw.id$respiratory.skipped_meals_binary)
```

Rename respiratory dataset joined to main dataset
```{r clean respiratory dataset}
coping.glad.respiratory <- coping.glad.respiratory.raw.id
